### 2013-06-17 context-dependence.tex
.

The common framework of the above examples is a one-dimensional grid of $n$ sites,
and each site is labeled with one of a finite set of states.
These change with time -- so, write $X_i(t)$ for the state of the $i^\mathrm{th}$ site at time $t$,
where each $X_i(t) \in \calS$, the set of possible states.
(To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
The time evolution is Markov, and is determined by a set of transition rates and associated patterns.  
``Patterns'' are sequences of states, and we let $|u|$ denote the length of the pattern $u$.
Transition rates are assumed to be local and homogeneous, so for example saying ``pattern $u$ changes to $v$ at rate $\mu$''
means that if $(X_i(t), \ldots, X_{i+\ell}(t)) = (u_1, \ldots, u_\ell)$ (where $\ell=|u|=|v|$),
then $\P\{(X_i(t+dt), \ldots, X_{i+\ell}(t+dt)) = (v_1, \ldots, v_\ell) = \mu dt + o(dt)$,
regardless of the location $i$ along the sequence.
We call such a $(\mu,u,v)$ a ``transition triple''.
The set of all transition triples $\calT = \{ (\mu_j,u_j,v_j) \}_{j=1}^{n_T}$ 
determines the time evolution of the process.

@@

\paragraph{The generator matrix}
For clarity, we should define how the set of transition rates determines the transition rate matrix,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate 
with which the process in state $x$ jumps to state $y$.
Let $x_i^{(\ell)} = (x_i, x_{i+1}, \ldots, x_{i+\ell-1})$ be the subsequence of length $\ell$ beginning at location $i$,
and for each $1\le i \le n$, and patterns $u,v \in \calS^\ell$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_j = y_j \quad &\text{for } j<i \\
  x_{i+k} = u_k \quad &\text{for } k < \ell \\
  y_{i+k} = v_k \quad &\text{for } k < \ell, and \\
  x_j = y_j \quad &\text{for } j\ge i+\ell ,
\end{cases}
\]
i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+\ell$, where $x$ matches with $u$ and $y$ matches with $v$.
Interpet these indices $\mod n$, even for short sequences, a point we will come back to.
We'll also need the set of transitions that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ j \st x \xrightarrow{i,u_j,v_j} y \}$.
Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
  G(n)_{x,y} = \sum_{i=1}^n \sum_{j \in J(i,x,y)}  \mu_j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
In principle, this gives us the transition probabilities for the process:
\begin{align} \label{eqn:full_likelihood}
  p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = (e^{tG_n})_{xy} .
\end{align}

>>>>

The common framework of the above examples is a one-dimensional grid of $n$ sites,
and each site is labeled with one of a finite set of states.
These change with time -- so, write $X_i(t)$ for the state of the $i^\mathrm{th}$ site at time $t$,
where each $X_i(t) \in \calS$, the set of possible states.
(To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
The time evolution is Markov, and is determined by a set of transition rates and associated patterns.  
``Patterns'' are sequences of states, and we let $|u|$ denote the length of the pattern $u$.
Transition rates are assumed to be local and homogeneous, so for example saying ``pattern $u$ changes to $v$ at rate $\mu$''
means that if $(X_i(t), \ldots, X_{i+\ell}(t)) = (u_1, \ldots, u_\ell)$ (where $\ell=|u|=|v|$),
then $\P\{(X_i(t+dt), \ldots, X_{i+\ell}(t+dt)) = (v_1, \ldots, v_\ell) = \mu dt + o(dt)$,
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
We call such a $(\mu,u,v)$ a ``transition triple''.
The set of all transition triples $\calT = \{ (\mu^j,u^j,v^j) \}_{j=1}^{n_T}$ 
determines the time evolution of the process.

@@

\paragraph{The generator matrix}
For clarity, we should define how the set of transition rates determines the transition rate matrix,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate 
with which the process in state $x$ jumps to state $y$.
% Let $x_i^{(\ell)} = (x_i, x_{i+1}, \ldots, x_{i+\ell-1})$ be the subsequence of length $\ell$ beginning at location $i$, and 
For each $1\le i \le n$, pattern length $\ell \ge 0$, and patterns $u,v \in \calS^\ell$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < \ell \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < \ell, and \\
  x_k = y_k \quad &\text{for } k\ge i+\ell ,
\end{cases}
\]
i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+\ell-1$, where $x$ matches with $u$ and $y$ matches with $v$.
(Interpet these indices modulo $n$, even for short sequences, a point we will come back to.)
We'll also need the set of defined transitions in $\calT$ 
that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
  G(n)_{x,y} = \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This is just to say that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
that could change $x$ into $y$.
In principle, this gives us the transition probabilities for the process:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = (e^{tG(n)})_{xy} .
\end{align}
